<!--
%\VignetteIndexEntry{Finding gendered words}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
-->

Finding gendered words
======================


```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = "", fig.path = "corpus-", fig.retina = TRUE)
options(width = 100)
```

In this vignette, we will look for gender-associated words  We will do this
by looking for words that follow the pronoun "she" more often than "he", or
vice-versa. The vignette is based on Julia Silge's blog post,
["Gender roles with text mining and n-grams"][silge-roles].


```{r}
library("corpus")

# RColorBrewer::brewer.pal(6, "Set1")
palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33"))

set.seed(0) # ensure consistent runs
```


We will be studying gender in the context of Jane Austen's completed novels,
downloaded from [Project Gutenberg][gutenberg].

```{r}
# Gutenberg IDs gotten from gutenbergr::gutenberg_works(author == "Austen, Jane")
(austen <- gutenberg_corpus(c(161, 1342, 141, 158, 121, 105)))
```

For tokens, we will be using the lower-cased non-punctuation words
```{r}
text_filter(austen)$drop_punct <- TRUE
```

Specifically, we will look at the bigrams from these novels where the first
type is a pronoun.

```{r}
pronouns <- c("he", "she")
bigram_counts <- term_stats(austen, ngrams = 2, types = TRUE,
                            subset = type1 %in% pronouns)
print(bigram_counts)
```

Above, we can see the most common bigrams with gendered pronouns as their first
types.


Next, we rearrange the data  into tabular form, with one row for each term and
one column for each pronoun:

```{r}
terms <-  with(bigram_counts,
               tapply(count, list(type2, type1), identity, default = 0))
head(terms)
```

It looks like some terms get emphasized in the text by underscores. We can
investigate some of these:

```{r}
# inspect a random sample of the occurrences
text_locate(austen, c("_did_", "_had", "_meant_"), random = TRUE)
```

Let's clean the terms by removing underscores, and recompute the counts:

```{r}
terms <-  with(bigram_counts,
               tapply(count, list(gsub("_", "", type2), type1),
                      sum, default = 0))
head(terms)
```

Here, we changed the aggregation function from `identity()` to `sum()` to sum
the counts over all types that are equal after removing underscores.


We want to find terms that are associated with particular gendered pronouns.
As an example, take term "loves": 

```{r}
term <- "loves"
i <- match(term, rownames(terms))
tab <- cbind(terms[i,], colSums(terms[-i,]))
colnames(tab) <- c(term, paste0("\u00ac", term))
print(tab)
```

When the first type in a bigram is "he", it is followed by "loves" in 5
instances and another type in 6988 instances. Here are the relative rates of
"loves", conditional on the first type in the bigram:

```{r}
(rates <- tab[,"loves"] / rowSums(tab))
```

The ratio of these rates is very close to one:

```{r}
rates[["she"]] / rates[["he"]]
```

The log (base-2) ratio of the rates is close to zero:
```{r}
log2(rates[["she"]]) - log2(rates[["he"]])
```

This log ratio is a good measure of the strength of association. When the
value is close to zero, there is no meaningful association. For large absolute
log ratios, we would say that there is a meaningful difference in usage
between the genders.

We compute the log ratios for all types. To avoid dividing by zero, and to
smooth the estimates when the counts are small, we add 0.5 to the count when
estimating the rate.

Some terms only appear once in the corpus. The rate estimates for these terms
will be unreliable, so we discard them.

```{r}
tot <- colSums(terms)
common_terms <- terms[rowSums(terms) > 1,]
he  <- (common_terms[,"he"] + 0.5) / (tot[["he"]] + 1)
she <- ((common_terms[,"she"] + 0.5) / (tot[["she"]] + 1))
log2_ratio <- log2(she) - log2(he)
```


Here are a histogram and normal probability plot of the estimates for the
remaining terms.

```{r fig.width = 12}
par(mfrow = c(1, 2))
hist(log2_ratio, breaks = "Scott", col = 2, border = "white")
qqnorm(log2_ratio, col = 2)
qqline(log2_ratio, col = 1, lty = 2)
```

This doesn't quite look like a normal, but the distribution is symmetric, and
the tails are about as heavy as those of a normal. So, the z-score will be a
reasonable measure of how typical or unusual each log rate is.

```{r}
z <- (log2_ratio - mean(log2_ratio)) / sd(log2_ratio)
```

Here are the words more than two standard deviations away from the mean log
ratio:

```{r}
sort(log2_ratio[abs(z) > 2])
```

The terms "remembered", "hurried", and "seized" have more "she" usages; the
terms "gravely", "believes", and "bowed" have more "he" usages.


It's hard to know which of these differences are meaningful without
quantifying the error associated with the estimates. Some words are common,
and we have reliable estimates of the log ratios. Other words are rare, and
the estimates are based on a small number of occurrences. In the rare case,
the estimates of the log ratios will be unreliable.


We need standard errors of our estimates. We can get these by starting with
the proportion estimates `he` and `she`, and then applying the
[delta method][delta-method].

The `he` and `she` rates are proportions. We get their standard errors by
using the usual formula for the standard error of a proportion, based on
the Binomial variance formula:

```{r}
he_se <- sqrt(he * (1 - he) / tot[["he"]])
she_se <- sqrt(he * (1 - he) / tot[["he"]])
```

To find the standard errors for the logarithms of these quantities, we use the
delta method. We multiply the standard error by the absolute value of the 
derivative of the logarithm function evaluated at the estimate:

```{r}
log2_he_se <- (1 / log(2)) * abs(1 / he) * he_se
log2_she_se <- (1 / log(2)) * abs(1 / she) * she_se
```

These formulas from using a Taylor expansion of `log2` around the estimate.

Finally, for the standard error of `log2_ratio`, we assume that the `log2_he`
and `log2_she` estimates are approximately independent, so that the variance
of their difference is the sum of their variances. This gives a formula for
the standard error of the log ratio:

```{r}
log2_ratio_se <- sqrt(log2_she_se^2 + log2_he_se^2)
```

```{r}
r <- rank(log2_ratio, ties.method = "first")

# significant cases
signif <- (abs(log2_ratio) / log2_ratio_se > 2)

i <- signif

xlim <- xlim <- range(r)
ylim <- range(log2_ratio,
              (log2_ratio - log2_ratio_se)[i],
              (log2_ratio + log2_ratio_se)[i])

plot(xlim, ylim, type = "n", xlab = "Rank",
     ylab = expression(paste(Log[2], " (She / He) Usage Rate")))
abline(h = 0, col = "gray", lty = 2)
points(r, log2_ratio, col = 2, cex = 0.5)

segments(r[i], (log2_ratio - log2_ratio_se)[i],
         r[i], (log2_ratio + log2_ratio_se)[i])
segments((r - 0.4)[i], (log2_ratio - log2_ratio_se)[i],
         (r + 0.4)[i], (log2_ratio - log2_ratio_se)[i])
segments((r - 0.4)[i], (log2_ratio + log2_ratio_se)[i],
         (r + 0.4)[i], (log2_ratio + log2_ratio_se)[i])
```


```{r fig.width = 12}

# order by log ratio, then name
o <- order(log2_ratio, names(log2_ratio))
est <- log2_ratio[o]
se <- log2_ratio_se[o]

# only keep significant cases
signif <- (abs(est) / se > 2)
est <- est[signif]
se <- se[signif]

# get the terms
term <- names(est)
i <- seq_along(term)

par(mar = c(5, 5, 2, 2) + 0.1)
xlim <- range(est - se, est + se)
ylim <- range(i)
plot(xlim, ylim, type = "n",
     xlab = "Relative appearance after 'she' compared to 'he'",
     ylab = "", axes = FALSE)

axis(2, at = i, labels = term, las = 1, cex.axis = 0.7)
abline(h = i, lty = 3, col = "gray")

ticks <- seq(-3, 6)
labels = paste0(2 ^ ticks, "x")
labels[ticks == 0] <- "Same"
axis(1, at = ticks, labels = labels)
abline(v = ticks, col = "gray", lwd = 0.5)

box()

col <- ifelse(est > 0, 3, 4)
segments(est - se, i, est + se, i, lwd = 1.5, col = col)
segments(est - se, i - 0.2, est - se, i + 0.2, lwd = 1.5, col = col)
segments(est + se, i - 0.2, est + se, i + 0.2, lwd = 1.5, col = col)
points(est, i, pch = 16, cex = 0.75, col = col)
```




[delta-method]: https://en.wikipedia.org/wiki/Delta_method
[gutenberg]: https://www.gutenberg.org/ "Project Gutenberg"
[silge-roles]: https://juliasilge.com/blog/gender-pronouns/ "Gender roles with text mining and n-grams"
